### 2017-03-24 paper/oopsla17-submission/evaluation-accuracy.tex
more eval

\paragraph{Our Classifiers}
We evaluate five classifiers, each trained on the full set of 276
features: 44 local syntactic features, 176 contextual syntactic
features, 55 typing features, and a single expression size feature.
%
\ES{should explain the make-up of these groups}
%
We preemptively discard expressions that are not part of the minimal
type error slice, we will explain the rationale for this in
\autoref{sec:feature-utility}.
%
Our classifiers are:
%
\begin{enumerate}
\item A logistic regression trained with a learning rate
  $\alpha = 0.001$, an L2 regularization rate $\lambda = 0.001$, and a
  mini-batch size of 200.
\item A decision tree \ES{HUMA: hyperparams?}.
\item A random forest with 30 estimators \ES{HUMA: hyperparams?}.
\item Two multi-layer perceptrons, both trained with $\alpha = 0.001$,
  $\lambda = 0.001$, and a mini-batch size of 200. The first MLP
  contains a single hidden layer of 10 neurons, and the second contains
  a hidden layer of 500 neurons. This allows us to investigate how well
  the MLP can \emph{compress} its model.
\end{enumerate}
%
All classifiers were trained for 20 iterations on one dataset before
being evaluated on the other.

@@

\paragraph{Results}
\autoref{fig:accuracy-results} shows the results of our experiment.
%
Localizing the type errors in our benchmarks amounted on average to
selecting one of 2--3 correct locations out of a slice of 10--11.
%
Our baseline of selecting at random is able to achieve 30\% Top 1
accuracy (58\% Top 3), while \ocaml achieves a Top 1 accuracy of 45\%.
%
Interestingly, one only needs two \emph{random} guesses to outperform
\ocaml, with 47\% accuracy.
%
\sherrloc outperforms both baselines with 56\% Top 1 accuracy (84\% Top
3), while \mycroft actually underperforms \ocaml with 38--41\% Top 1
accuracy.
%
Finally, we find that \emph{all} of our classifiers outperform \sherrloc,
ranging from 60--62\% Top 1 accuracy (85--88\% Top 3) for the \linear
classifier to 71--73\% Top 1 accuracy (91\% Top 3) for the \hiddenFH.

Interestingly there is little variation in accuracy between classifiers
with the exception of the \linear model, they all achieve around 70\%
Top 1 accuracy and around 90\% Top 3 accuracy.
%
In particular, the \hiddenT only loses around 2\% accuracy compared to
the \hiddenFH, which suggests that it can substantially compress the
model.
%
We also note that our classifiers consistently perform better when
trained on the \FALL programs and tested on the \SPRING programs than
vice versa, they appear to generalize better from the \FALL data.

>>>>

\paragraph{Our Classifiers}
We evaluate five classifiers, each trained on the full set of 276
features: 44 local syntactic features, 176 contextual syntactic
features, 55 typing features, and a single expression size feature.
%
\ES{should explain the make-up of these groups}
%
We preemptively discard expressions that are not part of the minimal
type error slice, we will explain the rationale for this in
\autoref{sec:feature-utility}.
%
Our classifiers are:
%
\begin{enumerate}
\item A logistic regression trained with a learning rate
  $\alpha = 0.001$, an L2 regularization rate $\lambda = 0.001$, and a
  mini-batch size of 200.
\item A decision tree \ES{HUMA: hyperparams?}.
\item A random forest with 30 estimators \ES{HUMA: hyperparams?}.
\item Two multi-layer perceptrons, both trained with $\alpha = 0.001$,
  $\lambda = 0.001$, and a mini-batch size of 200. The first MLP
  contains a single hidden layer of 10 neurons, and the second contains
  a hidden layer of 500 neurons. This allows us to investigate how well
  the MLP can \emph{compress} its model.
\end{enumerate}
%
All classifiers were trained for 20 epochs on one dataset
--- \ie they were shown each program 20 times ---
before being evaluated on the other.

@@

\paragraph{Results}
\autoref{fig:accuracy-results} shows the results of our experiment.
%
Localizing the type errors in our benchmarks amounted on average to
selecting one of 2--3 correct locations out of a slice of 10--11.
%
Our baseline of selecting at random is able to achieve 30\% Top-1
accuracy (58\% Top-3), while \ocaml achieves a Top-1 accuracy of 45\%.
%
Interestingly, one only needs two \emph{random} guesses to outperform
\ocaml, with 47\% accuracy.
%
\sherrloc outperforms both baselines with 56\% Top-1 accuracy (84\% Top
3), while \mycroft actually underperforms \ocaml with 38--41\% Top-1
accuracy.
%
Finally, we find that \emph{all} of our classifiers outperform \sherrloc,
ranging from 60--62\% Top-1 accuracy (85--88\% Top-3) for the \linear
classifier to 71--73\% Top-1 accuracy (91\% Top-3) for the \hiddenFH.

Interestingly there is little variation in accuracy between classifiers
with the exception of the \linear model, they all achieve around 70\%
Top-1 accuracy and around 90\% Top-3 accuracy.
%
In particular, the \hiddenT only loses around 2\% accuracy compared to
the \hiddenFH, which suggests that it can substantially compress the
model.
%
We also note that our classifiers consistently perform better when
trained on the \FALL programs and tested on the \SPRING programs than
vice versa, they appear to generalize better from the \FALL data.

### 2017-03-24 paper/oopsla17-submission/evaluation-utility.tex
more eval

\subsubsection{Type Error Slice}
\label{sec:type-error-slice}
The \InSlice feature should be highly predictive --- a fix must change
at least one expression in the type-error slice.
%
Thus, our first experiment seeks to quantify the impact of \InSlice by
comparing the accuracy of a linear model on three sets of features:
%
\begin{enumerate}
\item A baseline with only local syntactic features.
\item The local syntactic features of (1) extended with \InSlice.
\item The same features as (1), but we preemptively discard samples
  where \InSlice is \emph{disabled}.
\end{enumerate}
%
The key difference between (2) and (3) is that a classifier for (2) must
\emph{learn} that \InSlice is a strong predictor.
%
In contrast, a classifier for (3) must only learn about the syntactic
features, the decision to discard samples where \InSlice is disabled has
already been made by a human.
%
This has a few additional advantages: it reduces the set of candidate
locations by a factor of 7 on average, and it guarantees that any
prediction made by the classifier can fix the type error.
%
We expect that (2) will perform better than (1) as it contains more
information, and that (3) will perform better than (2) as the classifier
does not have to learn the importance of \InSlice.

We tested our hypothesis with a linear model cross-validated ($k=10$)
over the combined SP14/FA15 dataset. We used a learning rate
$\alpha=0.001$, L2 regularization rate $\lambda=0.001$, and mini-batch
size of 200. We trained for a single epoch on feature sets (1) and
(2), and for 8 epochs on (3), so that the total number of training samples
would be roughly equal for each feature set.

\begin{table}[ht]
  \centering
  \begin{tabular}{lrcrrrrcrrrr}
    \toprule
                 &             & & \multicolumn{4}{c} \linear        & & \multicolumn{4}{c} \hiddenF       \\
                                   \cmidrule{4-7}                        \cmidrule{9-12}
    Feature Set  & \# Features & & Top-1  & Top-2  & Top-3  & Recall & & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax & 47          & & 23.6\% & 42.6\% & 56.3\% & 19.6\% & & 30.9\% & 47.9\% & 58.6\% & 20.6\% \\
    + Slice      & 48          & & 46.4\% & 65.0\% & 75.4\% & 30.4\% & & 54.5\% & 70.5\% & 81.5\% & 34.7\% \\
    Only Slice   & 47          & & 54.9\% & 71.4\% & 82.5\% & 57.6\% & & 56.7\% & 72.3\% & 82.9\% & 58.0\% \\
    \bottomrule
  \end{tabular}
  \caption{
    Impact of type-error slice on accuracy.
    \ES{TODO: expand caption}
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
\end{table}

@@

Given the decisive benefits of filtering out expressions that do not
belong to the type-error slice, we will assume going forward that
\emph{all} data sets have been filtered.

@@

Next, we will investigate the relative impact of the other three classes
of features discussed in \autoref{sec:features}.
%
For this we consider again a baseline of only local syntactic features,
extended by each combination of
%
(1) expression size,
(2) contextual syntactic features, and
(3) typing features.
%
As before we perform a 10-fold cross-validation with a learning rate and
L2 regularization rate of 0.001 and a mini-batch size of 200, but we
train for a full 10 epochs.

\begin{table}[ht]
  \centering
  \begin{tabular}{lrcrrrrcrrrr}
    \toprule
                             &             & & \multicolumn{4}{c} \linear        & & \multicolumn{4}{c} \hiddenFH      \\
                                               \cmidrule{4-7}                        \cmidrule{9-12}
    Feature Set              & \# Features & & Top-1  & Top-2  & Top-3  & Recall & & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax             &  44         & & 55.0\% & 71.6\% & 82.6\% & 57.7\% & & 56.3\% & 72.1\% & 82.8\% & 57.6\% \\
    \midrule
    + Size                   &  45         & & 55.8\% & 72.8\% & 82.5\% & 57.3\% & & 60.5\% & 75.0\% & 83.8\% & 57.9\% \\
    + Context                & 220         & & 59.9\% & 77.7\% & 86.4\% & 63.0\% & & 71.4\% & 84.1\% & 90.8\% & 69.5\% \\
    + Types                  & 102         & & 62.2\% & 77.7\% & 85.7\% & 62.2\% & & 73.2\% & 84.8\% & 90.2\% & 69.2\% \\
    \midrule
    + Context + Size         & 221         & & 60.2\% & 77.9\% & 86.0\% & 62.5\% & & 71.7\% & 83.5\% & 90.7\% & 69.3\% \\
    + Types + Size           & 103         & & 62.0\% & 78.2\% & 85.6\% & 62.3\% & & 73.5\% & 85.8\% & 91.4\% & 70.5\% \\
    + Context + Types        & 275         & & 63.2\% & 80.3\% & 87.9\% & 65.4\% & & 77.3\% & 87.5\% & 92.4\% & 72.9\% \\
    \midrule
    + Context + Types + Size & 276         & & 62.3\% & 80.0\% & 88.0\% & 65.4\% & & 77.2\% & 87.9\% & 92.7\% & 72.9\% \\
    \bottomrule
  \end{tabular}
  % \begin{minipage}{0.49\linewidth}
  % \centering
  % \hiddenF
  % \begin{tabular}{lrrrr}
  %   \toprule
  %   Feature Set                 & Top-1  & Top-2  & Top-3  & Recall \\
  %   \midrule
  %   Local Syntax                & 56.9\% & 72.2\% & 82.8\% & 57.9\% \\
  %   \midrule
  %   + Size                      & 59.7\% & 74.6\% & 83.0\% & 57.4\% \\
  %   + Context                   & 70.9\% & 83.7\% & 90.4\% & 69.2\% \\
  %   + Types                     & 72.1\% & 84.1\% & 90.3\% & 69.3\% \\
  %   \midrule
  %   + Size + Context            & 69.8\% & 83.5\% & 90.2\% & 68.6\% \\
  %   + Size + Types              & 72.3\% & 84.6\% & 90.3\% & 69.5\% \\
  %   + Context + Types           & 75.5\% & 86.4\% & 91.5\% & 71.7\% \\
  %   \midrule
  %   + All                       & 75.0\% & 86.8\% & 91.9\% & 72.0\% \\
  %   \bottomrule
  % \end{tabular}
  % \end{minipage}
  % \\
  % \begin{minipage}{0.49\linewidth}
  % \centering
  % \hiddenFH
  % \begin{tabular}{lrrrr}
  %   \toprule
  %   Feature Set                 \\
  %   \midrule
  %   Local Syntax                \\
  %   \midrule
  %   + Size                      \\
  %   + Context                   \\
  %   + Types                     \\
  %   \midrule
  %   + Size + Context            \\
  %   + Size + Types              \\
  %   + Context + Types           \\
  %   \midrule
  %   + All                       \\
  %   \bottomrule
  % \end{tabular}
  % \end{minipage}
  \caption{
    Impact of contextual features on accuracy.
    \ES{TODO: expand caption}
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
\end{table}

>>>>

\subsubsection{Type Error Slice}
\label{sec:type-error-slice}
The \InSlice feature should be highly predictive --- a fix must change
at least one expression in the type-error slice.
%
Thus, our first experiment seeks to quantify the impact of \InSlice by
comparing the accuracy of a linear model on three sets of features:
%
\begin{enumerate}
\item A baseline with only local syntactic features.
\item The features of (1) extended with \InSlice.
\item The same features as (1), but we preemptively discard samples
  where \InSlice is \emph{disabled}.
\end{enumerate}
%
The key difference between (2) and (3) is that a classifier for (2) must
\emph{learn} that \InSlice is a strong predictor.
%
In contrast, a classifier for (3) must only learn about the syntactic
features, the decision to discard samples where \InSlice is disabled has
already been made by a human.
%
This has a few additional advantages: it reduces the set of candidate
locations by a factor of 7 on average, and it guarantees that any
prediction made by the classifier can fix the type error.
%
We expect that (2) will perform better than (1) as it contains more
information, and that (3) will perform better than (2) as the classifier
does not have to learn the importance of \InSlice.

We tested our hypothesis with a linear model cross-validated ($k=10$)
over the combined SP14/FA15 dataset. We used a learning rate
$\alpha=0.001$, L2 regularization rate $\lambda=0.001$, and mini-batch
size of 200. We trained for a single epoch on feature sets (1) and
(2), and for 8 epochs on (3), so that the total number of training samples
would be roughly equal for each feature set.
\ES{introduce the MLP too}
\ES{use MLP-500 instead?}
\begin{table}[ht]
  \centering
  \begin{tabular}{lrcrrrrcrrrr}
    \toprule
                       &             & & \multicolumn{4}{c} \linear        & & \multicolumn{4}{c} \hiddenF       \\
                                         \cmidrule{4-7}                        \cmidrule{9-12}
    Feature Set        & \# Features & & Top-1  & Top-2  & Top-3  & Recall & & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax       & 47          & & 23.6\% & 42.6\% & 56.3\% & 19.6\% & & 30.9\% & 47.9\% & 58.6\% & 20.6\% \\
    + \InSlice         & 48          & & 46.4\% & 65.0\% & 75.4\% & 30.4\% & & 54.5\% & 70.5\% & 81.5\% & 34.7\% \\
    Filter by \InSlice & 47          & & 54.9\% & 71.4\% & 82.5\% & 57.6\% & & 56.7\% & 72.3\% & 82.9\% & 58.0\% \\
    \bottomrule
  \end{tabular}
  \caption{
    Impact of type-error slice on accuracy.
    \ES{TODO: expand caption}
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
\end{table}

@@

Given the decisive benefits of filtering out expressions that do not
belong to the type-error slice, we choose to filter all programs by
\InSlice.

@@

Next, we will investigate the relative impact of the other three classes
of features discussed in \autoref{sec:features}.
%
For this we consider again a baseline of only local syntactic features,
extended by each combination of
%
(1) expression size,
(2) contextual syntactic features, and
(3) typing features.
%
As before we perform a 10-fold cross-validation with $\alpha = 0.001$,
$\lambda = 0.001$, and a mini-batch size of 200, but we
train for a full 20 epochs.

\begin{table}[ht]
  \centering
  \begin{tabular}{lrcrrrrcrrrr}
    \toprule
                             &             & & \multicolumn{4}{c} \linear        & & \multicolumn{4}{c} \hiddenFH      \\
                                               \cmidrule{4-7}                        \cmidrule{9-12}
    Feature Set              & \# Features & & Top-1  & Top-2  & Top-3  & Recall & & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax             &  44         & & 55.0\% & 71.6\% & 82.6\% & 57.7\% & & 56.3\% & 72.1\% & 82.8\% & 57.6\% \\
    \midrule
    + Size                   &  45         & & 55.8\% & 72.8\% & 82.5\% & 57.3\% & & 60.5\% & 75.0\% & 83.8\% & 57.9\% \\
    + Context                & 220         & & 59.9\% & 77.7\% & 86.4\% & 63.0\% & & 71.4\% & 84.1\% & 90.8\% & 69.5\% \\
    + Types                  & 102         & & 62.2\% & 77.7\% & 85.7\% & 62.2\% & & 73.2\% & 84.8\% & 90.2\% & 69.2\% \\
    \midrule
    + Context + Size         & 221         & & 60.2\% & 77.9\% & 86.0\% & 62.5\% & & 71.7\% & 83.5\% & 90.7\% & 69.3\% \\
    + Types + Size           & 103         & & 62.0\% & 78.2\% & 85.6\% & 62.3\% & & 73.5\% & 85.8\% & 91.4\% & 70.5\% \\
    + Context + Types        & 275         & & 63.2\% & 80.3\% & 87.9\% & 65.4\% & & 77.3\% & 87.5\% & 92.4\% & 72.9\% \\
    \midrule
    + Context + Types + Size & 276         & & 62.3\% & 80.0\% & 88.0\% & 65.4\% & & 77.2\% & 87.9\% & 92.7\% & 72.9\% \\
    \bottomrule
  \end{tabular}
  % \begin{minipage}{0.49\linewidth}
  % \centering
  % \hiddenF
  % \begin{tabular}{lrrrr}
  %   \toprule
  %   Feature Set                 & Top-1  & Top-2  & Top-3  & Recall \\
  %   \midrule
  %   Local Syntax                & 56.9\% & 72.2\% & 82.8\% & 57.9\% \\
  %   \midrule
  %   + Size                      & 59.7\% & 74.6\% & 83.0\% & 57.4\% \\
  %   + Context                   & 70.9\% & 83.7\% & 90.4\% & 69.2\% \\
  %   + Types                     & 72.1\% & 84.1\% & 90.3\% & 69.3\% \\
  %   \midrule
  %   + Size + Context            & 69.8\% & 83.5\% & 90.2\% & 68.6\% \\
  %   + Size + Types              & 72.3\% & 84.6\% & 90.3\% & 69.5\% \\
  %   + Context + Types           & 75.5\% & 86.4\% & 91.5\% & 71.7\% \\
  %   \midrule
  %   + All                       & 75.0\% & 86.8\% & 91.9\% & 72.0\% \\
  %   \bottomrule
  % \end{tabular}
  % \end{minipage}
  \caption{
    Impact of contextual features on accuracy.
    \ES{TODO: expand caption}
    \ES{TODO: load these numbers from CSV}
  }\label{tab:contextual-features}
\end{table}

\autoref{tab:contextual-features} summarizes the results of our experiment.
%
As we can see, the \linear classifier and the \hiddenFH start off
competitive when given only local syntactic features, but the \hiddenFH
quickly begins to outperform as more features are added.
%
\ES{HEREHEREHERHE}

### 2017-03-24 paper/oopsla17-submission/introduction.tex
more eval

\subsection{Contributions}
\label{sec:contributions}
Thus, we identify the following novel contributions of our work.
%
\begin{enumerate}
\item An experimental evaluation of state-of-the-art type-error
  localization techniques based not on expert \emph{opinion}, but
  on the eventual \emph{fix} implemented by a novice user.
  %
  \ES{need to explain \emph{why} this is superior..}
\item A machine learning approach to \emph{modeling} (novice) type
  errors that can outperform the state-of-the-art by 10--30\% in
  localizing errors.
  %
  Our model can be trained on a relatively small amount of data --- we
  used a single class of around 50 students --- and appears to
  generalize to other instances of the same class.
  %
  \ES{and hopefully completely separate classes!}
  %
  In contrast to many approaches to \emph{fault localization} from the
  software engineering community, our model does not use any linguistic
  modeling techniques, \eg \emph{n-grams} over the token stream; rather,
  it relies entirely on features of the abstract syntax tree and a
  partial typing derivation.
  %
  \ES{if there's time, investigate adding Wes' n-gram model}
\end{enumerate}

>>>>

\subsection{Contributions}
\label{sec:contributions}
Thus, we identify the following novel contributions of our work.
%
\begin{enumerate}
\item An large-scale evaluation of state-of-the-art type error
  localization techniques on over 4,500 student programs.
  %
  We collected a time-series of \ocaml programs from our students, and
  use their subsequent \emph{fixes} to ill-typed programs as
  \emph{oracles} for the location of the error.
  %
  This is the most extensive evaluation of the accuracy of type error
  localization techniques that we are aware of.
  % based not on expert \emph{opinion}, but
  % on the eventual \emph{fix} implemented by a novice user.
  %
  % \ES{need to explain \emph{why} this is superior..}
\item A machine learning approach to \emph{modeling} (novice) type
  errors that can outperform the state-of-the-art by 15--30\% in
  localizing errors.
  %
  Our classifiers can be trained on a relatively small amount of data --- we
  used a single class of around 50 students --- and appear to
  generalize to other instances of the same class.
  %
  \ES{and hopefully completely separate classes!}
  %
  In contrast to many approaches to \emph{fault localization} from the
  software engineering community, our model does not use any linguistic
  modeling techniques, \eg \emph{n-grams} over the token stream; rather,
  it relies entirely on features of the abstract syntax tree and a
  partial typing derivation.
  %
  \ES{if there's time, investigate adding Wes' n-gram model}
\end{enumerate}
